---
title: "PA1_template"
author: "Rajeshwaran Gunalan"
date: "7 July 2018"
output: html_document
---

#### Loading the necessary libraries 

``` {r , echo=TRUE , message=FALSE}
library(caret)
library(ggplot2)
library(plyr)
library(dplyr)
library(stringr)
```

<br>
<br>


### Loading and preprocessing the data


#### Loading the data to memory and doing the necessary format changes  

```{r , echo=TRUE , message=FALSE}

Activity <- read.csv("C:/Data Science/John Hopkins - Coursera/Reproducible Research/Assignment 1/activity.csv")
Activity$date <- as.character(Activity$date)
Activity$date <- as.factor(Activity$date)

```


<br>
<br>


#### What is mean total number of steps taken per day?


```{r  , echo=TRUE}
TotalStepsPerDay <- Activity %>% group_by(date) %>% summarise(TotalSteps = sum(steps))
head(TotalStepsPerDay , 10)

```

<br>
<br>


#### Histogram of steps taken per day 


```{r  , echo=TRUE}

hist(TotalStepsPerDay$TotalSteps)

```


#### Mean and Median of total steps per day


```{r  , echo=TRUE}

meanTotalSteps <- mean(TotalStepsPerDay$TotalSteps , na.rm = TRUE)
meanTotalSteps

medianTotalSteps <- median(TotalStepsPerDay$TotalSteps , na.rm = TRUE)
medianTotalSteps
```

<br>
<br>


### What is the average daily activity pattern?


#### Time series plot

```{r  , echo=TRUE}

Activity$interval <- as.factor(Activity$interval)

IntervalStepsMean <- Activity %>% group_by(interval) %>% summarise(MeanSteps=mean(steps , na.rm = TRUE))

plot(as.numeric( IntervalStepsMean$interval)  , IntervalStepsMean$MeanSteps , type="l")

```

<br>
<br>

#### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

``` {r , echo=TRUE}

MostActiveInterval <- IntervalStepsMean[order(IntervalStepsMean$MeanSteps , decreasing = TRUE),]$interval [1]

as.numeric(as.character(MostActiveInterval))

```

<br>
<br>

### Imputing Missing Values 

<br>

#### Calculate and report the total number of missing values in the dataset 

``` {r , echo=TRUE}

nrow(Activity[ which(is.na(Activity$steps)) , ])


``` 

<br>
<br>

#### Filling in the missing values and creating a new dataset that is equal to the original dataset but with the missing data filled in.

```{r , echo=TRUE}


MissingValueIndices <- which(is.na(Activity$steps))


Activity_MissingImputed <- Activity


for( i in MissingValueIndices)
{
  
  Activity_MissingImputed[i , "steps"] <- inner_join(Activity[i, ] , IntervalStepsMean , by = 'interval')$MeanSteps
  
  
}


```


<br>
<br>

### Histogram of total number of steps each day after imputation 


```{r , echo=TRUE}

TotalStepsPerDay_Imputed <- Activity_MissingImputed %>% group_by(date) %>% summarise(TotalSteps = sum(steps))

hist(TotalStepsPerDay_Imputed$TotalSteps)

meanTotalSteps_Imputed <- mean(TotalStepsPerDay_Imputed$TotalSteps , na.rm = TRUE)

medianTotalSteps_Imputed <- median(TotalStepsPerDay_Imputed$TotalSteps , na.rm = TRUE)


```



#### Mean and Median of total steps each day after imputation 


``` {r , echo=TRUE}

meanTotalSteps_Imputed

medianTotalSteps_Imputed

```


#### Impact of Mean and Median after imputation 


```{r}

meanTotalSteps - meanTotalSteps_Imputed

medianTotalSteps - medianTotalSteps_Imputed

```

#### Since I had used the interval wise mean for imputing , there is no impact on the mean. Howeever, there is a small difference in the median 

<br>
<br>




### Are there differences in activity patterns between weekdays and weekends?


#### Creating a new feature indicating Weekday or Weekend 


```{r , echo=TRUE}

Activity_MissingImputed$date <- as.character(Activity_MissingImputed$date)

Activity_MissingImputed$date <- as.Date(Activity_MissingImputed$date)

Activity_MissingImputed$DayOfWeek <- weekdays(Activity_MissingImputed$date )


Activity_MissingImputed$WeekendFlag <- NA

Activity_MissingImputed$WeekendFlag <- as.character(Activity_MissingImputed$WeekendFlag)

Activity_MissingImputed[which(Activity_MissingImputed$DayOfWeek == "Saturday") , ]$WeekendFlag <- "weekend"

Activity_MissingImputed[which(Activity_MissingImputed$DayOfWeek == "Sunday") , ]$WeekendFlag <- "weekend"

Activity_MissingImputed[which(is.na(Activity_MissingImputed$WeekendFlag)) , ]$WeekendFlag <- "weekday"

Activity_MissingImputed$WeekendFlag <- as.factor(Activity_MissingImputed$WeekendFlag)


```

#### Plot to check if there is an difference in activity levels  between Weekday and weekend


```{r , echo=TRUE}


IntervalStepsMean_WeekDay <- Activity_MissingImputed %>% group_by(WeekendFlag , interval) %>% summarise(MeanSteps=mean(steps , na.rm = TRUE))


ggplot(data=IntervalStepsMean_WeekDay , mapping = aes(as.numeric(interval) , MeanSteps)) + geom_line()+
                                            facet_wrap( ~ WeekendFlag)


```




